Finiteness in Cubical Type Theory

### 2020-06-01 sections/finiteness-predicates.tex
rearranged more

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power,
so any closure proof we have on one can be transferred to the other.
In particular, it means that manifest Bishop finiteness is closed under
\(\Sigma\).
\subsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{B}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\mathcal{B})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \begin{multicols}{2}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
  \end{multicols}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

It turns out the answer is no!
\begin{lemma}
  \begin{equation}
    \neg (\Sigma(A : \mathbf{Type}) , \mathcal{C}(A) \times \neg \mathcal{B}(A))
  \end{equation}
\end{lemma}
\begin{proof}
  We will actually prove a slightly more general statement.
  For any type \(A\), the following holds:
  \begin{equation}
    \neg (\lVert A \rVert \times \neg A)
  \end{equation}
  The solution becomes more clear if we write out the definition of \(\neg\):
  \begin{equation}
    \frac{\lVert A \rVert \;\;\; A \rightarrow \bot }{\bot}
  \end{equation}
  We clearly need to apply a function of type \(A \rightarrow \bot\) to a value
  of type \(\lVert A \rVert\).
  Luckily, this is permissible, as \(\bot\) is a mere proposition.
\end{proof}
\section{Manifest Enumerability}
Since we have now discovered that all three predicates defined so far have equal
restrictiveness, the clear next step is to

So far, we have pushed finiteness predicates in the direction of making them
``less informative'': we have not, however,

@@

Secondly, it means the predicate does not imply decidable equality.
More significantly, it allows the predicate to be defined over non-set types,
like the circle.
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignedat}{3}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
      | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignedat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\section{Kuratowski Finiteness}
The first thing we must define is a representation of subsets.
\begin{definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

@@

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{lemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{lemma}
\begin{lemma}
  This circle \(S^1\) is Kuratowski finite.
\end{lemma}
\section{Relations Between Each Finiteness Definition} \label{relations}
We will now look at the arrows in figure~\ref{finite-classification}.
\section{Split Enumerability and Manifest Bishop Finiteness}
\section{Split Enumerability and Manifest Enumerability}
\begin{lemma} \label{split-enum-to-manifest-enum}
  Any split enumerable type is manifestly enumerable.
\end{lemma}
This lemma is proven by truncating the membership proof in split enumerability.

\begin{lemma} \label{manifest-enum-to-split-enum}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{lemma}

\section{Manifest Bishop Finiteness and Cardinal Finiteness}
\begin{lemma} \label{manifest-bishop-to-cardinal}
  Any manifest Bishop finite type is cardinal finite.
\end{lemma}
\begin{theorem} \label{cardinal-to-manifest-bishop}
  Any cardinal finite type with a total order is Bishop finite.
\end{theorem}
The proof for this particular theorem is quite involved in the formalisation, so
we only give its sketch here.
First, note that we actually convert to manifest enumerability first: this can
be converted to split enumerability with decidable equality, which is provided
by cardinal finiteness.

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\section{Cardinal Finiteness and Kuratowski Finiteness}
\begin{lemma} \label{cardinal-kuratowski}
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}^f(A) \times \text{Discrete}(A)
  \end{equation}
\end{lemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\mathcal{K}^f(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions.
The actual functions themselves are proven in our formalisation, as well as in
\cite{fruminFiniteSetsHomotopy2018}.

>>>>

We have now proved that every manifestly Bishop finite type is split enumerable,
and vice versa.
While the types are not \emph{equivalent} (there are more split enumerable
proofs than there are manifest Bishop finite proofs), they are of equal power,
so any closure proof we have on one can be transferred to the other.
In particular, it means that manifest Bishop finiteness is closed under
\(\Sigma\).
\subsection{From Manifest Bishop Finiteness to Equivalence}
We have seen that split enumerability was in fact a split-surjection in
disguise.
We will now see that manifest Bishop finiteness is in fact an \emph{equivalence}
in disguise.
\emph{equivalence} with \(\mathbf{Fin}\).
\begin{lemma} \label{bishop-equiv}
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  \begin{align*}
     \mathcal{B}(A) &
    \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , x \inunique \mathit{xs}
    && \text{def.~\ref{bish-def} }(\mathcal{B})
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(x \in \mathit{xs})
    && \text{eqn.~\ref{uniq-memb-def} } (\inunique)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \Pi {(x : A)} , \text{isContr}(\text{fib}_{\text{snd}(\mathit{xs})}(x))
    && \text{eqn.~\ref{container-membership} } (\in)
    \\
    & \simeq \Sigma (\mathit{xs} : \textbf{List}(A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{is-equiv-def} (isEquiv)}
    \\
    & \simeq \Sigma (\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket (A)) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{def.~\ref{List} } (\mathbf{List})
    \\
    & \simeq \Sigma (\mathit{xs} : \Sigma (n : \mathbb{N}) , \Pi (i : \mathbf{Fin}(n)) , A) , \text{isEquiv}(\text{snd}(\mathit{xs}))
    && \text{eqn.~\ref{container-interp} } (\llbracket \wc \rrbracket)
    \\
    & \simeq \Sigma (n : \mathbb{N}) , \Sigma (f : \mathbf{Fin}(n) \rightarrow A) , \text{isEquiv}(f)
    && \text{Reassociation of } \Sigma
    \\
    & \simeq \Sigma (n : \mathbb{N}) , ( \mathbf{Fin}(n) \simeq A )
    && \text{eqn.~\ref{equiv-def} } (\simeq) \; \qedhere
  \end{align*}
\end{proof}
This proof is almost identical\footnotemark to the proof for
lemma~\ref{split-enum-is-split-surj}: it reveals that
enumeration-based finiteness predicates are simply another perspective on
relation-based ones.
\footnotetext{
  Unfortunately in our formalisation this proof cannot be a single line: for
  performance reasons \(\simeq\) is defined as a record type with eta-equality
  disabled, instead of the definition here which uses \(\Sigma\).
}

As we are working in CuTT, a proof of equivalence between two types gives us the
ability to \emph{transport} proofs from one type to the other.
This is extremely powerful, as we will see.
\subsection{Closure Under \(\Pi\)}
The glaring omission from our closure proofs under type formers so far has been
the \(\Pi\) type: we have not proved closure under functions, dependent or
otherwise.
In MLTT, this is of course not provable: since all of the finiteness predicates
we have seen so far imply decidable equality, and since we don't have any kind
of decidable equality on functions in MLTT, we know that we won't be able to
show that any kind of function is finite; even one like \(\mathbf{Bool}
\rightarrow \mathbf{Bool}\).

CuTT is not so restricted.
Since we have things like function extensionality and transport, we can indeed
prove the finiteness of function types.
Our proof here makes use directly of the univalence axiom, and makes use
furthermore of all the previous closure proofs.
We will prove this closure on split enumerability, rather than on manifest
Bishop finiteness, as it requires slightly less legwork in the proof itself, but
of course we can derive the proof on manifest Bishop finiteness in a few lines.
\begin{lemma} \label{split-enum-pi}
  Split enumerability is closed under dependent functions.
  (\(\Pi\)-types).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi {(x : A)} , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a split enumerable type, and \(U\) be a type family from \(A\),
  which is split enumerable over all points of \(A\).

  As \(A\) is split enumerable, we know that it is also manifestly Bishop finite
  (lemma~\ref{split-enum-to-manifest-bishop}), and consequently we know \(A
  \simeq \mathbf{Fin}(n)\), for some \(n\) (lemma~\ref{bishop-equiv}).
  We can therefore replace all occurrences of \(A\) with \(\mathbf{Fin}(n)\),
  changing our goal to:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi (x : \mathbf{Fin}(n)) , U(x)\right)
    }
  \end{equation}

  We then define the type of \(n\)-tuples over some type family \(T :
  \mathbf{Fin}(n) \rightarrow \mathbf{Type}\).
  \begin{equation}
    \begin{alignedat}{3}
      & \mathbf{Tuple}(0, T)   &&\coloneqq \top \\
      & \mathbf{Tuple}(n+1, T) &&\coloneqq T(0) \times \mathbf{Tuple}(n, T \circ \text{suc})
    \end{alignedat}
  \end{equation}
  We can show that this type is equivalent to functions (proven in our formalisation):
  \begin{equation}
    \Pi(x : \mathbf{Fin}(n)) , U(x) \simeq \mathbf{Tuple}(n, U)
  \end{equation}
  And therefore we can simplify again our goal to the following:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\mathbf{Tuple}(n, U)\right)
    }
  \end{equation}

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}
\section{Cardinal Finiteness}
While we have removed some of the unnecessary information from our finiteness
predicates, one piece still remains.
\begin{agdalisting}
  The two following proofs are both valid proofs of the finiteness of
  \(\mathbf{Bool}\), and both do not include any duplicates.
  However they still differ:
  \begin{multicols}{2}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-inst}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool-rev}
  \end{multicols}
\end{agdalisting}
Each finiteness predicate so far has contained an \emph{ordering} of the
underlying type.
For our purposes, this is too much information: it means that when constructing
the ``category of finite sets'' later on, instead of each type having one
canonical representative, it will have \(n!\), where \(n\) is the cardinality of
the type\footnotemark.

@@

It turns out the answer is no!
\begin{lemma}
  \begin{equation}
    \neg (\Sigma(A : \mathbf{Type}) , \mathcal{C}(A) \times \neg \mathcal{B}(A))
  \end{equation}
\end{lemma}
\begin{proof}
  We will actually prove a slightly more general statement.
  For any type \(A\), the following holds:
  \begin{equation}
    \neg (\lVert A \rVert \times \neg A)
  \end{equation}
  The solution becomes more clear if we write out the definition of \(\neg\):
  \begin{equation}
    \frac{\lVert A \rVert \;\;\; A \rightarrow \bot }{\bot}
  \end{equation}
  We clearly need to apply a function of type \(A \rightarrow \bot\) to a value
  of type \(\lVert A \rVert\).
  Luckily, this is permissible, as \(\bot\) is a mere proposition.
\end{proof}

\subsection{Going from Cardinal Finiteness to Manifest Bishop Finiteness}
\begin{lemma} \label{manifest-bishop-to-cardinal}
  Any manifest Bishop finite type is cardinal finite.
\end{lemma}
\begin{theorem} \label{cardinal-to-manifest-bishop}
  Any cardinal finite type with a total order is Bishop finite.
\end{theorem}
The proof for this particular theorem is quite involved in the formalisation, so
we only give its sketch here.
First, note that we actually convert to manifest enumerability first: this can
be converted to split enumerability with decidable equality, which is provided
by cardinal finiteness.

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\subsection{Closure}
Since we don't have a function of type \(\mathcal{C}(A) \rightarrow
\mathcal{B}(A)\), closure proofs on \(\mathcal{B}\) do not transfer over to
\(\mathcal{C}\) trivially (unlike with \(\mathcal{E}!\) and \(\mathcal{B}\)).
The cases for \(\bot\), \(\top\), and \(\mathbf{Bool}\) are simple to adapt: we
can just propositionally truncate their Bishop finiteness proof.

Non-dependent operators like \(\times\), \(\uplus\), and \(\rightarrow\) are
also relatively straightforward: since \(\lVert {\wc} \rVert\) forms a monad, we
can apply \(n\)-ary functions to values inside it, combining them together.
\begin{agdalisting}
  The fact that \(\lVert \wc \rVert\) forms a monad means that we can lift
  \(n\)-ary functions like the following:
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{times-clos-sig}
  Into a truncated context:
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{times-clos-impl}
\end{agdalisting}

Unfortunately, for the dependent type formers like \(\Sigma\) and \(\Pi\), the
same trick does not work.
We have closure proofs like:
\begin{equation}
  \frac{
    \mathcal{B}(A) \; \; \; \Pi(x : A) , \mathcal{B}(U(x))
  }{
    \mathcal{B}(\Pi \; A \; U)
  }
\end{equation}
If we apply the monadic truncation trick we can derive closure proofs like the
following:
\begin{equation}
  \frac{
    \lVert \mathcal{B}(A) \rVert \; \; \; \lVert \Pi(x : A) , \mathcal{B}(U(x)) \rVert
  }{
    \lVert \mathcal{B}(\Pi \; A \; U) \rVert
  }
\end{equation}
However our \emph{desired} closure proof is the following:
\begin{equation}
  \frac{
    \lVert \mathcal{B}(A) \rVert \; \; \; \Pi(x : A) , \lVert \mathcal{B}(U(x)) \rVert
  }{
    \lVert \mathcal{B}(\Pi \; A \; U) \rVert
  }
\end{equation}
They don't match!

The solution would be to find a function of the following type:
\begin{equation}
  (\Pi(x : A) , \lVert \mathcal{B}(U(x)) \rVert) \rightarrow
  \lVert \Pi(x : A) , \mathcal{B}(U(x)) \rVert
\end{equation}
However we might be disheartened at realising that this is a required goal: the
above equation is \emph{extremely} similar to the axiom of choice!
\begin{definition}[Axiom of Choice]
  In HoTT, the axiom of choice is commonly defined as follows \cite[lemma
  3.8.2]{hottbook}.
  For any set \(A\), and a type family \(U\) which is a set at all the points
  of \(A\), the following function exists:
  \begin{equation}
    \left( \Pi(x : A) ,  \lVert U(x) \rVert \right) \rightarrow \lVert \Pi(x : A) , U(x) \rVert
  \end{equation}
\end{definition}
Luckily the axiom of choice \emph{does} hold for cardinally finite types,
allowing us to prove the following:
\begin{lemma}
  \begin{equation}
    \mathcal{C}(A) \rightarrow (\Pi(x : A) , \lVert U(x) \rVert) \rightarrow \lVert \Pi(x : A) , U(x) \rVert
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a cardinally finite type, \(U\) be a type family on \(A\), and
  \(f\) be a dependent function of type \(\Pi(x : A) , \lVert U(x) \rVert\).

  First, since our goal is itself propositionally truncated, we have access to
  values under truncations: put another way, in the context of proving our goal,
  we can rely on the fact that \(A\) is manifestly Bishop finite.
  Using the same technique as we did in lemma~\ref{split-enum-pi}, we can switch
  from working with dependent functions from \(A\) to \(n\)-tuples, where \(n\)
  is the cardinality of \(A\).
  This changes our goal to the following:
  \begin{equation}
    \mathbf{Tuple}(n, \lVert \wc \rVert \circ U) \rightarrow \lVert \mathbf{Tuple}(n, U) \rVert
  \end{equation}
  Since \(\lVert \wc \rVert\) is closed under finite products, this function
  exists (in fact, using the fact that \(\lVert \wc \rVert\) forms a monad, we
  can recognise this function as \verb+sequenceA+ from the \verb+Traversable+
  class in Haskell).
\end{proof}

This gets us all of the necessary closure proofs on \(\mathcal{C}\).
\section{Manifest Enumerability}
We have now explored quite far in the ``less informative'' direction.
However, all three predicates we have examined are equally \emph{restrictive}:
in this section we will see a predicate which is much less restrictive.
In particular, this predicate ranges over non-set types.

@@

Secondly, it means the predicate does not imply decidable equality.
More significantly, it allows the predicate to be defined over non-set types,
like the circle.
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{equation}
    \begin{alignedat}{3}
      S^1 \coloneqq & \; \text{base} &&: S^1 ; \\
      | & \; \text{loop} &&: \text{base} \equiv \text{base} ; 
    \end{alignedat}
  \end{equation}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}
\begin{lemma}
  The circle \(S^1\) is manifestly enumerable.
\end{lemma}
\subsection{Relation To Split Enumerability }
\begin{lemma} \label{split-enum-to-manifest-enum}
  Any split enumerable type is manifestly enumerable.
\end{lemma}
This lemma is proven by truncating the membership proof in split enumerability.

\begin{lemma} \label{manifest-enum-to-split-enum}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{lemma}
\section{Kuratowski Finiteness}
The first thing we must define is a representation of subsets.
\begin{definition}[Kuratowski Finite Subset]
  \(\mathcal{K}(A)\) is the type of Kuratowski-finite subsets of \(A\).
  \begin{inductivetype}{\mathcal{K}(A)}
    \inductivetypeclause{\left[ \right]}{\mathcal{K}(A)}
    \inductivetypeclause{\wc \dblcolon \wc}{A \rightarrow \mathcal{K}(A) \rightarrow \mathcal{K}(A)}
    \inductivetypeclause{\text{com}}{\Pi (x, y: A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon y \dblcolon \mathit{xs} \equiv y \dblcolon x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{dup}}{\Pi (x : A) , \Pi (\mathit{xs} : \mathcal{K}(A)) , x \dblcolon x \dblcolon \mathit{xs} \equiv x \dblcolon \mathit{xs}}
    \inductivetypeclause{\text{trunc}}{\Pi (\mathit{xs}, \mathit{ys}: \mathcal{K}(A)) , \Pi (p, q : \mathit{xs} \equiv \mathit{ys}) , p \equiv q}
  \end{inductivetype}
  We define it as a HIT (definition~\ref{HITs}).
  The first two constructors are point constructors, giving ways to create
  values of type \(\mathcal{K}(A)\).
  They are also recognisable as the two constructors for finite lists, a type
  which represents the free monoid.

@@

We also have the following two lemmas, proven in both
\cite{fruminFiniteSetsHomotopy2018} and our formalisation.
\begin{lemma}
  \(\mathcal{K}^f\) is a mere proposition.
\end{lemma}
\begin{lemma}
  This circle \(S^1\) is Kuratowski finite.
\end{lemma}
\subsection{Relation to Cardinal Finiteness}
\begin{lemma} \label{cardinal-kuratowski}
  \begin{equation}
    \mathcal{C}(A) \simeq \mathcal{K}^f(A) \times \text{Discrete}(A)
  \end{equation}
\end{lemma}
This proof is constructed by providing a pair of functions: one from
\(\mathcal{C}(A)\) to \(\mathcal{K}^f(A) \times \text{Discrete}(A)\), and one the
other way.
This pair implies an equivalence, because both source and target are
propositions.
The actual functions themselves are proven in our formalisation, as well as in
\cite{fruminFiniteSetsHomotopy2018}.

### 2020-06-01 sections/topos.tex
now compiles without errors
rearranged more

\begin{lemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are split enumerable.
\end{lemma}
\begin{lemma} \label{split-enum-sigma}
  Split enumerability is closed under \(\Sigma\).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi(x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma(x : A) , U(x))
    }
  \end{equation}
\end{lemma}
From this we can derive split enumerability of non-dependent sums and products,
as both are definable in terms of \(\Sigma\).

\begin{lemma}
  Split enumerability is closed under dependent functions.
  (\(\Pi\)-types).
  \begin{equation}
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi {(x : A)} , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{lemma}
\begin{proof}
  Let \(A\) be a split enumerable type, and \(U\) be a type family from \(A\),
  which is split enumerable over all points of \(A\).

  As \(A\) is split enumerable, we know that it is also manifestly Bishop finite
  (lemma~\ref{split-enum-to-manifest-bishop}), and consequently we know \(A
  \simeq \mathbf{Fin}(n)\), for some \(n\) (lemma~\ref{bishop-equiv}).
  We can therefore replace all occurrences of \(A\) with \(\mathbf{Fin}(n)\),
  changing our goal to:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\Pi (x : \mathbf{Fin}(n)) , U(x)\right)
    }
  \end{equation}

  We then define the type of \(n\)-tuples over some type family \(T :
  \mathbf{Fin}(n) \rightarrow \mathbf{Type}\).
  \begin{equation}
    \begin{alignedat}{3}
      & \mathbf{Tuple}(0, T)   &&\coloneqq \top \\
      & \mathbf{Tuple}(n+1, T) &&\coloneqq T(0) \times \mathbf{Tuple}(n, T \circ \text{suc})
    \end{alignedat}
  \end{equation}
  We can show that this type is equivalent to functions (proven in our formalisation):
  \begin{equation}
    \Pi(x : \mathbf{Fin}(n)) , U(x) \simeq \mathbf{Tuple}(n, U)
  \end{equation}
  And therefore we can simplify again our goal to the following:
  \begin{equation}
    \frac{
      \mathcal{E}!(\mathbf{Fin}(n)) \; \; \; \Pi (x : \mathbf{Fin}(n)) , \mathcal{E}!\left( U(x) \right)
    }{
      \mathcal{E}!\left(\mathbf{Tuple}(n, U)\right)
    }
  \end{equation}

  We can prove this goal by showing that \(\mathbf{Tuple}(n, U)\) is split
  enumerable: it is made up of finitely many products of points of \(U\), which
  are themselves split enumerable, and \(\top\), which is also split enumerable.
  Lemma~\ref{split-enum-sigma} shows us that the product of finitely many split
  enumerable types is itself split enumerable, proving our goal.
\end{proof}

Lifting the above closure proofs to work on proofs of \(\mathcal{C}\) is not
necessarily straightforward.
\(\lVert \wc \rVert\) forms a monad, giving us access to a powerful syntax for
combining proofs.
From this, we can derive closure under \(\bot\), \(\top\), \(\mathbf{Bool}\),
\(\times\), \(\uplus\), and \(\rightarrow\).
The dependent case is not so straightforward.
We need the following lemma, a version of the axiom of choice on finite sets,
to prove closure under \(\Sigma\) and \(\Pi\).
\begin{lemma}
  \begin{equation}
    \mathcal{C}(A) \rightarrow (\Pi(x : A) , \lVert U(x) \rVert) \rightarrow \lVert \Pi(x : A) , U(x) \rVert
  \end{equation}
\end{lemma}
\section{The Category of Finite Sets}
HoTT and CuTT seem to be especially suitable settings for formalisations of
category theory.
The univalence axiom in particular allows us to treat categorical isomorphisms
as equalities, saving us from the dreaded ``setoid hell''.

>>>>

% \begin{lemma}
%   \(\bot\), \(\top\), and \(\mathbf{Bool}\) are split enumerable.
% \end{lemma}
% \begin{lemma} \label{split-enum-sigma}
%   Split enumerability is closed under \(\Sigma\).
%   \begin{equation}
%     \frac{
%       \mathcal{E}!(A) \; \; \; \Pi(x : A) , \mathcal{E}!(U(x))
%     }{
%       \mathcal{E}!(\Sigma(x : A) , U(x))
%     }
%   \end{equation}
% \end{lemma}
% From this we can derive split enumerability of non-dependent sums and products,
% as both are definable in terms of \(\Sigma\).

\section{The Category of Finite Sets}
HoTT and CuTT seem to be especially suitable settings for formalisations of
category theory.
The univalence axiom in particular allows us to treat categorical isomorphisms
as equalities, saving us from the dreaded ``setoid hell''.
